Community Analysis

### 2022-07-26 R/08-community-analysis.Rmd
update analysis

s5 <-
  rare_data %>% 
  filter(passage == "G5") %>% 
  cor.test(~ ch4 + rare, data = ., method = "spearman", alternative = "two.sided")

min_seqs <- paste(substr(as.character(min_seqs), 1, 3), substr(as.character(min_seqs), 4, 6), sep = ",")
save(kt, wt, rare_median, s, min_seqs, file = "Output/richness_models.RData")

@@

## Beta diversity test passage by treatment

```{r permanova-passage-treatment, warnings=FALSE}

dist_df <-
  rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

@@

betadisper(rare_dist_bray, meta_dist$passage_treat) %>% 
  permutest()

@@

# Distance-based RDA Methane

```{r full-dbrda}

meta_dist %>% 
  select(passage_treat, ch4) %>% 
  mutate(log_ch4 = log_methane(ch4)) %>% 
  ggplot(aes(x = passage_treat, y = log_ch4)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Passage x Treatment", y = "Log10 Methane")

dbrda_ch4 <- dbrda(rare_dist_bray ~ rank(ch4), 
                      data = meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

dbrda_ch4 %>% 
  permutest(permutations = 9999)

percent_explained <- 100 * eigenvals(dbrda_ch4) / sum(eigenvals(dbrda_ch4))

@@

@@

```{r pre-dbrda-data-subsets}

# aes_data %>% 
#   filter(treat == "P") %>% 
#   group_by(asv) %>% 
#   mutate(total = sum(count)) %>% 
#   filter(total != 0) %>% 
#   ungroup() %>% 
#   select(-total) %>% 
#   group_by(sample) %>% 
#   summarize(n_seqs = sum(count)) %>% 
#   summarize(min = min(n_seqs))

treatP_df <-
  aes_data %>%  
  filter(treat == "P") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

rownames(treatP_df) <- treatP_df$sample
treatP_df <- treatP_df[, -1]

# aes_data %>% 
#   filter(passage == "G5") %>% 
#   group_by(asv) %>% 
#   mutate(total = sum(count)) %>% 
#   filter(total != 0) %>% 
#   ungroup() %>% 
#   select(-total) %>% 
#   group_by(sample) %>% 
#   summarize(n_seqs = sum(count)) %>% 
#   summarize(min = min(n_seqs))

passage2_df <-
  aes_data %>%  
  filter(passage == "G2") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

rownames(passage2_df) <- passage2_df$sample
passage2_df <- passage2_df[, -1]

passage5_df <-
  aes_data %>%  
  filter(passage == "G5") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

```{r treat-dbrda}

if(file.exists("Output/treat_rare_dist_bray.rds")) {
  treat_rare_dist_bray <- 
    readRDS("Output/treat_rare_dist_bray.rds")
} else {
treat_rare_dist_bray <- avgdist(treatP_df, dmethod="bray", sample=seq_range[1])
  treat_rare_dist_bray %>% 
    saveRDS("Output/treat_rare_dist_bray.rds")
}

treat_dist_df <-
  treat_rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

treat_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(treat_dist_df, by = "sample")

treat_dbrda_ch4 <- dbrda(treat_rare_dist_bray ~ rank(ch4), 
                      data = treat_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

treat_dbrda_ch4 %>% 
  permutest(permutations = 9999)

percent_explained <- 100 * eigenvals(treat_dbrda_ch4) / sum(eigenvals(treat_dbrda_ch4))

treat_dbrda_scores <-
  scores(treat_dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

treat_dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(treat_dbrda_scores, by = "sample")

treat_dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Positive'),
                       values = c('darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

```{r passage2-dbrda}
if(file.exists("Output/passage2_rare_dist_bray.rds")) {
  passage2_rare_dist_bray <- 
    readRDS("Output/passage2_rare_dist_bray.rds")
} else {
passage2_rare_dist_bray <- avgdist(passage2_df, dmethod="bray", sample=seq_range[1])
  passage2_rare_dist_bray %>% 
    saveRDS("Output/passage2_rare_dist_bray.rds")
}

passage2_dist_df <-
  passage2_rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

passage2_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage2_dist_df, by = "sample")

passage2_dbrda_ch4 <- dbrda(passage2_rare_dist_bray ~ rank(ch4), 
                      data = passage2_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

passage2_dbrda_permtest <- 
passage2_dbrda_ch4 %>% 
  permutest(permutations = 9999)

passage2_dbrda_permtest <- 
passage2_dbrda_ch4 %>% 
  anova(permutations = 9999) %>% 
  tidy()

passage2_percent_explained <- 100 * eigenvals(passage2_dbrda_ch4) / sum(eigenvals(passage2_dbrda_ch4))

passage2_dbrda_scores <-
  scores(passage2_dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

passage2_dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage2_dbrda_scores, by = "sample")

passage2_dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

```{r passage5-dbrda}
if(file.exists("Output/passage5_rare_dist_bray.rds")) {
  passage5_rare_dist_bray <- 
    readRDS("Output/passage5_rare_dist_bray.rds")
} else {
passage5_rare_dist_bray <- avgdist(passage5_df, dmethod="bray", sample=seq_range[1])
  passage5_rare_dist_bray %>% 
    saveRDS("Output/passage5_rare_dist_bray.rds")
}

passage5_dist_df <-
  passage5_rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

passage5_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage5_dist_df, by = "sample")

passage5_dbrda_ch4 <- dbrda(passage5_rare_dist_bray ~ rank(ch4), 
                      data = passage5_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

passage5_dbrda_permtest <-
passage5_dbrda_ch4 %>% 
  anova(permutations = 9999) %>% 
  tidy

passage5_percent_explained <- 100 * eigenvals(passage5_dbrda_ch4) / sum(eigenvals(passage5_dbrda_ch4))

passage5_dbrda_scores <-
  scores(passage5_dbrda_ch4, display = "sites") %>% 
  as_tibble(rownames = "sample")

passage5_dbrda_data <-
  aes_data %>% 
  select(sample, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(passage5_dbrda_scores, by = "sample")

passage5_dbrda_data %>%   
  ggplot(aes(x = dbRDA1, y = MDS1, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage") +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) +
  theme_bw() +
  theme(
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

```{r}
save(passage2_dbrda_permtest, 
     passage5_dbrda_permtest, 
     passage5_percent_explained,
     file = "Output/dbrda.RData")
```

The overall dbRDA is significant and rank of methane explains 5% of the variation.
Subsetting the Positive treatment in both G2 and G5, it is not significant: no
correlation between dissimilarity and rank of methane between G2 and G5. This
makes sense, because most of that variation would be due to selection in the jar.
However, subsetting passage G5 and comparing N and P, 10% of dissimilarity is
explained by rank of CH4.

@@

```{r methanotroph-lists}

methanotrophs <- c('Methylococcaceae', 
                   'Methylothermaceae', 
                   'Methylocystaceae', 
                   'Beijerinckiaceae', 
                   'Methylacidiphilaceae')
methanotrophs <- c(methanotrophs, paste(methanotrophs, "unclassified", sep = "_"))

@@

@@

```{r stop}

>>>>

s5 <-
  rare_data %>% 
  filter(passage == "G5") %>% 
  cor.test(~ ch4 + rare, data = ., method = "spearman", alternative = "two.sided") %>%
  tidy

min_seqs <- paste(substr(as.character(min_seqs), 1, 3), substr(as.character(min_seqs), 4, 6), sep = ",")
save(kt, wt, rare_median, s5, min_seqs, file = "Output/richness_models.RData")

@@

```{r}
dist_df <-
  rare_dist_bray %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

@@

betadisper(rare_dist_bray, meta_dist$passage_treat) %>% 
  anova()

@@

## dbrda methane in passage 5

```{r dbrda, warnings=FALSE}
passage5_df <-
  aes_data %>%  
  filter(passage == "G5") %>% 
  group_by(asv) %>% 
  mutate(total = sum(count)) %>% 
  filter(total != 0) %>% 
  ungroup() %>% 
  select(-total) %>% 
  select(sample, asv, count) %>%
  pivot_wider(names_from = "asv", values_from = "count", values_fill = 0) %>% 
  as.data.frame()

if(file.exists("Output/rare_dist_bray_p5.rds")) {
  rare_dist_bray_p5 <- 
    readRDS("Output/rare_dist_bray_p5.rds")
} else {
rare_dist_bray_p5 <- avgdist(passage5_df, dmethod="bray", sample=seq_range[1])
  rare_dist_bray_p5 %>% 
    saveRDS("Output/rare_dist_bray_p5.rds")
}

rare_dist_df <-
  rare_dist_bray_p5 %>% 
  as.matrix %>% 
  as_tibble(rownames = "sample")

dbrda_meta_dist <-
  aes_data %>% 
  select(sample, passage_treat, passage, treat, ch4) %>% 
  distinct %>% 
  inner_join(rare_dist_df, by = "sample")

dbrda_ch4 <- dbrda(rare_dist_bray_p5 ~ rank(ch4), 
                      data = dbrda_meta_dist, 
                      na.action = "na.omit",
                      add = TRUE)

dbrda_aov <-
  dbrda_ch4 %>% 
  anova()

dbrda_percent_explained <- 100 * eigenvals(dbrda_ch4) / sum(eigenvals(dbrda_ch4))

@@

save(dbrda_aov, dbrda_percent_explained, file = "Output/dbrda.RData")

@@

@@

```{r methanotroph-lists}
# List of methanotrophs from Table 1 of Stein 2012 
troph_genera <- c('Methylococcus',
                  'Methylocaldum',
                  'Methylobacter',
                  'Methylomicrobium',
                  'Methylomonas',
                  'Methylosarcina',
                  'Methylosoma',
                  'Crenothrixa',
                  'Clonothrixa',
                  'Methylosphaera',
                  'Methylohalobiusa',
                  'Methylothermusa',
                  'Methylocystis',
                  'Methylosinus',
                  'Methylocella',
                  'Methylocapsa',
                  'Methyloferula',
                  'Methylomirabilis',
                  'Methylacidiphilum')
troph_genera <- c(troph_genera, paste(troph_genera, "unclassified", sep = "_"))

methanotrophs <- c('Methylococcaceae', 
                   'Methylothermaceae', 
                   'Methylocystaceae', 
                   'Beijerinckiaceae', 
                   'NC10',
                   'Methylacidiphilaceae')
methanotrophs <- c(methanotrophs, paste(methanotrophs, "unclassified", sep = "_"))

@@

troph_asvs <-
  aes_data %>% 
  select(Kingdom:Species, asv) %>% 
  distinct() %>% 
  filter(Genus %in% troph_genera) %$%
  asv

@@
